\section{Discussion and outlook}
  \label{sec:discussion}

  \subsection{Summary}
    \label{subsec:summary}

    We have shown that the projective spectral entropy functional $\Spi[g]$ admits
    a natural local thermodynamic structure.
    The diagonal heat kernel defines a local spectral energy density $u(x;t)$ whose
    Seeley--DeWitt expansion yields a local multiplier field
    $\beta(x)^{-1} = \beta_0^{-1} - \frac{1}{6}R(x) + O(\nabla^2 R)$
    decomposed into a universal UV part and a geometric curvature correction.

    The local spectral first law $\delta\Spi = \int\beta^{-1}\delta E_\Pi$ is
    established as a compatibility theorem, not a definition: both sides are
    independently defined from spectral and matter data respectively.

    The spectral equilibrium principle --- extremizing $\Spi$ at fixed projected
    matter content --- yields Einstein's equation $G_{\mu\nu} = 8\pi G_N T_{\mu\nu}$
    as an Euler--Lagrange condition.
    The factor $8\pi G_N$ is fixed by the product of two universal normalizations from
    the geometric and matter sides, with no free parameter.

  \subsection{Limitations}
    \label{subsec:limitations}

    \paragraph{Euclidean setting.}
      The entire construction is performed on Riemannian manifolds.
      The extension to Lorentzian signature requires analytic continuation of the heat
      kernel, control over the retarded Green function, and a definition of $u(x;t)$ in
      the hyperbolic setting.
      This is the content of the Lorentzian completion program described in~\cite{Gravity20}.

    \paragraph{Static foliation.}
      The projected matter energy $E_\Pi[\Sigma]$ is defined on a fixed spatial slice.
      A fully covariant formulation would require a foliation-independent definition,
      likely via the covariant phase space formalism.

    \paragraph{Higher Seeley--DeWitt corrections.}
      The equilibrium condition~\eqref{eq:einstein} holds at leading IR order.
      In the regime $R\ell_\chi^2 \sim 1$, higher Seeley--DeWitt contributions (from $a_4$
      and beyond) become comparable, and the field equation receives quadratic curvature
      corrections of the form $G_{\mu\nu} + \ell_\chi^2 B_{\mu\nu} + \cdots =
8\pi G_N T_{\mu\nu}$.

    \paragraph{Matter coupling.}
      The projected stress tensor $T^{(\Pi)}_{\mu\nu}$ is defined formally from $W_\Pi$.
      An explicit construction of $W_\Pi$ in terms of standard matter fields requires
      identifying the admissible projected sector for each matter species.
      This is addressed in the Cosmochrony framework~\cite{Cosmochrony} but not derived
      here.

\subsection{Outlook}
  \label{subsec:outlook}

  The present framework opens several directions.
  First, the Lorentzian completion of Paper~3 will test whether the equilibrium
  condition survives analytic continuation and whether it reproduces linearized
  gravitational wave propagation.
  Second, the higher Seeley--DeWitt regime ($a_4$ dominant, $\xi = 1/6$) may
  provide a spectral derivation of higher-curvature gravity theories.
  Third, the connection between the local field $\beta(x)^{-1}$ and the local
  Unruh effect could be made precise if a Lorentzian horizon is introduced as a
  special case of the general framework.
