\section{Introduction}
  \label{sec:intro}

  In a companion paper~\cite{Gravity20}, we established that the projective spectral
  entropy functional
  \begin{equation}
    \Spi[g] \equiv \half\log\det' \Ag ,
    \label{eq:Spi-def}
  \end{equation}
  defined by a minimal Laplace-type operator $\Ag = -\nabla_g^2$ on a four-dimensional
  Riemannian manifold $(M,g)$, admits a well-defined renormalized metric variation.
  In the infrared-dominant regime where all dimensionless curvature invariants satisfy
  $R\ell_\chi^2 \ll 1$, the renormalized spectral stress tensor decomposes as
  \begin{equation}
    \frac{\delta\Spi^{\mathrm ren}}{\delta g^{\mu\nu}}
    = c_{\mathrm EH}({\mu})\,G_{\mu\nu} + c_\Lambda(\mu)\,g_{\mu\nu}
    + \beta_{\mathrm loc}\,B_{\mu\nu} + T^{\mathrm non\text{-}loc}_{\mu\nu} + A_{\mu\nu},
    \label{eq:decomp-paper1}
  \end{equation}
  with $c_{\mathrm EH} = (16\pi G_N)^{-1}$ identified via induced-gravity renormalization;
  higher-derivative terms are suppressed by powers of $\ell_\chi^2$.

  The goal of the present paper is to develop the local thermodynamic interpretation
  of this structure.
  We show that the diagonal heat kernel $\hk(x,x;t)$ defines a natural local spectral
  energy density $u(x;t)$, whose curvature expansion produces a local multiplier field
  $\beta(x)^{-1}$ with a universal part and a geometric correction proportional to
  the scalar curvature $R(x)$.
  This field plays the role of a local inverse temperature, defined purely from spectral
  data, without thermodynamic input.

  We then show that the compatibility between the geometric and matter sides of the
  metric variation is precisely the condition
  \begin{equation}
    \delta\Spi = \int_\Sigma \beta(x)^{-1}\,\delta E_\Pi(x)\,\dd^3 x ,
    \label{eq:first-law-intro}
  \end{equation}
  which we call the \emph{local spectral first law}.
  This is not a postulate but a structural consequence of the Seeley--DeWitt
  hierarchy.

  Finally, we impose a spectral equilibrium principle: admissible geometries are
  those for which $\Spi$ is stationary at fixed projected matter content.
  The Euler--Lagrange equation of this constrained extremum is Einstein's field
  equation, obtained without any reference to local horizons, Rindler observers,
  or causal structure.

  \medskip
  \noindent\textbf{Relation to Jacobson's derivation.}
  Jacobson~\cite{Jacobson1995} derived Einstein's equation by imposing the Clausius
  relation $\delta S = \delta Q/T$ on all local Rindler horizons, using the
  Bekenstein--Hawking entropy proportional to area.
  The present derivation is structurally parallel but differs in three essential ways:
  (i) the entropy functional is $\Spi$, not an area functional;
  (ii) the temperature field $\beta(x)^{-1}$ is defined from spectral data, not from
  Rindler kinematics;
  (iii) the field equation arises from a variational extremum, not from a horizon
  equilibrium condition.
  The relation to Jacobson's derivation is discussed further in
  Section~\ref{sec:relation-jacobson}.

  \medskip
  \noindent\textbf{Structure of the paper.}
  Section~\ref{sec:spectral-locality} develops the local spectral energy density and
  the multiplier field $\beta(x)$.
  Section~\ref{sec:first-law} establishes the local spectral first law as a
  compatibility theorem.
  Section~\ref{sec:equilibrium} formulates the spectral equilibrium principle and
  derives the Einstein equation.
  Section~\ref{sec:normalization} verifies the normalization of the factor $8\pi G$.
  Section~\ref{sec:relation-jacobson} discusses the relation to previous thermodynamic
  derivations.
  Section~\ref{sec:discussion} collects limitations and outlook.
