\section{Relation to previous thermodynamic derivations}
  \label{sec:relation-jacobson}

  \subsection{Jacobson's derivation}
    \label{subsec:jacobson}

    Jacobson~\cite{Jacobson1995} derived Einstein's equation by applying the
    Clausius relation $\delta S = \delta Q / T$ to all local Rindler horizons.
    The key inputs are:
    (i) entropy proportional to horizon area, $S = A/(4G)$;
    (ii) temperature given by the Unruh formula, $T = a/(2\pi)$;
    (iii) heat flow $\delta Q = \int T_{\mu\nu}\chi^\mu\dd\Sigma^\nu$ across the horizon;
    (iv) the Raychaudhuri equation linking area change to matter flux.

    The present derivation parallels Jacobson's logical structure --- entropy, temperature,
    energy, equilibrium condition --- but replaces each ingredient with a spectral
    counterpart:
    (i) entropy is $\Spi = \frac{1}{2}\log\det' A_g$ (not an area functional);
    (ii) temperature is the local spectral multiplier $\beta(x)^{-1}$ (not Rindler);
    (iii) energy is the projected matter content $E_\Pi$ (not a horizon heat flux);
    (iv) equilibrium is a variational extremum (not a horizon balance condition).

    As a consequence, the present derivation is independent of the causal structure of
    spacetime: it holds on Riemannian manifolds and does not require a Lorentzian
    horizon.
    The extension to Lorentzian signature is discussed in Section~\ref{sec:discussion}.

  \subsection{Analogy with local Unruh temperature}
    \label{subsec:unruh-analogy}

    In the Unruh effect, the local temperature experienced by a Rindler observer with
    proper acceleration $a$ is $T_{\mathrm Unruh} = a/(2\pi)$.
    This is kinematic: it depends on the observer's trajectory.

    The local multiplier field $\beta(x)^{-1} = \beta_0^{-1} - \frac{1}{6}R(x)$
    has a formally analogous structure: a universal part $\beta_0^{-1}$ plus a local
    correction.
    However, the analogy is purely formal.
    The correction $-\frac{1}{6}R(x)$ is geometric (curvature-driven), not kinematic
    (acceleration-driven).
    No horizon is present or implied.
    The analogy should not be taken to mean that $\beta(x)^{-1}$ is a physical
    temperature in the Unruh sense: it is a spectral multiplier defined from
    heat-kernel data, not from a quantum state on a wedge.

  \subsection{Wald's Noether charge entropy}
    \label{subsec:wald}

    Wald~\cite{Wald1993} showed that black hole entropy for higher-derivative theories
    of gravity is given by a Noether charge, not by the area.
    The present framework is consistent with Wald's result in the following sense:
    the spectral entropy $\Spi$ is defined from the full operator $\Ag$, which encodes
    higher-derivative geometric information through the $a_4$ and higher Seeley--DeWitt
    coefficients.
    In the UV regime (or for $\xi \neq 0$), the dominant contribution to $\Spi$ shifts
    from the $a_2$-derived Einstein term to the $a_4$-derived quadratic curvature terms,
    whose variation would give a higher-derivative stress tensor consistent with Wald's
    formula.
    A detailed comparison lies beyond the scope of the present paper.
