\section{Spectral equilibrium principle and Einstein equation}
  \label{sec:equilibrium}

  \subsection{Formulation of the equilibrium principle}
    \label{subsec:equilibrium-formulation}

    We now impose the following variational principle.

    \begin{definition}[Spectral equilibrium]
      \label{def:equilibrium}
      An admissible geometry $g$ is in \emph{spectral equilibrium} if it is a
      stationary point of the functional
      \begin{equation}
        \mathcal{F}[g,\psi]
        \equiv
        \Spi[g] - \beta_*^{-1}\,W_\Pi[g,\psi]
        \label{eq:F-functional}
      \end{equation}
      under arbitrary compactly supported metric variations $\delta g^{\mu\nu}$,
      with $\beta_*^{-1}$ a constant Lagrange multiplier.
    \end{definition}

    The functional $\mathcal{F}$ is the spectral analogue of a free energy: it extremizes
    entropic gain against the projected matter cost.
    The Lagrange multiplier $\beta_*^{-1}$ enforces the constraint that the projected
    matter content $W_\Pi$ is held fixed along admissible deformations.

    \begin{remark}[Nature of $\beta_*^{-1}$]
      \label{rem:beta-star}
      The global multiplier $\beta_*^{-1}$ in~\eqref{eq:F-functional} is distinct from
      the local field $\beta(x)^{-1}$ defined in Section~\ref{sec:spectral-locality}.
      The local field encodes spatial curvature variation; the global multiplier encodes
      the overall scale of the matter-geometry coupling.
      Their relation is discussed in Section~\ref{sec:normalization}.
    \end{remark}

  \subsection{Euler--Lagrange equation}
    \label{subsec:euler-lagrange}

    \begin{theorem}[Einstein equation as spectral equilibrium]
      \label{thm:einstein}
      In the infrared-dominant regime $R\ell_\chi^2 \ll 1$, the Euler--Lagrange
      equation of the spectral equilibrium condition $\delta\mathcal{F} = 0$ is
      \begin{equation}
        G_{\mu\nu} = 8\pi G_N\,T^{(\Pi)}_{\mu\nu},
        \label{eq:einstein}
      \end{equation}
      up to a cosmological term and corrections suppressed by $O(R\ell_\chi^2)$.
    \end{theorem}

    \begin{proof}
      Varying $\mathcal{F}[g,\psi]$ with respect to $g^{\mu\nu}$:
      \begin{equation}
        \delta\mathcal{F}
        = \delta\Spi - \beta_*^{-1}\,\delta W_\Pi = 0.
        \label{eq:delta-F}
      \end{equation}
      Substituting~\eqref{eq:deltaS-EH} and~\eqref{eq:W-variation}:
      \begin{equation}
        \frac{1}{16\pi G_N}\int_M\sqrt{g}\;G_{\mu\nu}\,\delta g^{\mu\nu}
        = \frac{\beta_*^{-1}}{2}\int_M\sqrt{g}\;T^{(\Pi)}_{\mu\nu}\,\delta g^{\mu\nu}
        + O(R\ell_\chi^2).
        \label{eq:EL-integrated}
      \end{equation}
      Since this holds for arbitrary compactly supported $\delta g^{\mu\nu}$, the
      fundamental lemma of the calculus of variations gives pointwise:
      \begin{equation}
        \frac{1}{16\pi G_N}\,G_{\mu\nu}
        = \frac{\beta_*^{-1}}{2}\,T^{(\Pi)}_{\mu\nu}
        + O(R\ell_\chi^2).
        \label{eq:EL-pointwise}
      \end{equation}
      For $\beta_*^{-1} = 1$ (see Section~\ref{sec:normalization}), this gives
      \begin{equation}
        G_{\mu\nu} = 8\pi G_N\,T^{(\Pi)}_{\mu\nu} + O(R\ell_\chi^2).
        \label{eq:einstein-proof}
      \end{equation}
    \end{proof}

    \begin{remark}[Absence of causal assumptions]
      \label{rem:no-horizon}
      The derivation requires no horizon structure, no local Rindler observers,
      no Raychaudhuri equation, and no assumption on causal structure.
      The field equation emerges from a purely variational condition on the spectral
      functional $\mathcal{F}$, defined on Riemannian manifolds.
      Its extension to Lorentzian signature is a separate question, discussed in
      Section~\ref{sec:discussion}.
    \end{remark}
