\section{Local spectral first law as a compatibility theorem}
  \label{sec:first-law}

  \subsection{Projected matter energy}
    \label{subsec:projected-matter}

    We define the projected matter energy on a spatial slice $\Sigma$ independently
    of the spectral multiplier field.
    Let $W_\Pi[g,\psi]$ be the effective matter action associated with the admissible
    projected sector.
    Its metric variation defines the projected stress tensor $T^{(\Pi)}_{\mu\nu}$ via
    the standard covariant identity
    \begin{equation}
      \delta W_\Pi[g,\psi]
      = \half\int_M \dd^4x\,\sqrt{g}\;T^{(\Pi)}_{\mu\nu}\,\delta g^{\mu\nu}.
      \label{eq:W-variation}
    \end{equation}
    On a chosen spatial slice $\Sigma$ with future-pointing unit normal $n^\mu$ and
    local Killing vector $\xi^\mu$, the projected matter energy is defined by
    \begin{equation}
      E_\Pi[\Sigma]
      \equiv
      \int_\Sigma T^{(\Pi)}_{\mu\nu}\,\xi^\mu n^\nu\,\dd\Sigma,
      \label{eq:E-Pi-def}
    \end{equation}
    and its metric variation is
    \begin{equation}
      \delta E_\Pi
      = \half\int_\Sigma T^{(\Pi)}_{\mu\nu}\,\delta g^{\mu\nu}\,\dd\Sigma.
      \label{eq:delta-E-Pi}
    \end{equation}
    This definition is independent of $\beta(x)$ and of the heat-kernel structure.

  \subsection{Local spectral first law}
    \label{subsec:first-law}

    \begin{proposition}[Local spectral first law --- compatibility condition]
      \label{thm:first-law}
      In the infrared regime $R\ell_\chi^2 \ll 1$, and at leading order in the
      Seeley--DeWitt derivative expansion, the metric variation of the renormalized
      spectral entropy satisfies
      \begin{equation}
        \delta\Spi^{\mathrm ren}
        = \int_\Sigma \beta(x)^{-1}\,\delta E_\Pi(x)\,\dd^3 x
        \label{eq:first-law}
      \end{equation}
      as a structural compatibility identity between the geometric and matter sides
      of the variational problem.
      This identity holds if and only if the geometry satisfies the Einstein equation.
      It is therefore an equivalence condition, not an independent postulate.
    \end{proposition}

    \begin{proof}[Proof (compatibility argument)]
        From~\cite{Gravity20}, the infrared-dominant local part of $\delta\Spi^{\mathrm ren}$
        contains the Einstein tensor:
        \begin{equation}
          \delta\Spi^{\mathrm ren}
          \supset
          \frac{1}{16\pi G_N}
          \int_M \dd^4x\,\sqrt{g}\;G_{\mu\nu}\,\delta g^{\mu\nu}
          + O(R^2\ell_\chi^2).
          \label{eq:deltaS-EH}
        \end{equation}
        The Bianchi identity $\nabla^\mu G_{\mu\nu} = 0$ implies that the integrand
        $G_{\mu\nu}\delta g^{\mu\nu}$ admits a local density representation on $\Sigma$
        once $\delta g^{\mu\nu}$ is supported in a compact region.
        On the matter side, from~\eqref{eq:delta-E-Pi}:
        \begin{equation}
          \delta E_\Pi
          = \half\int_\Sigma T^{(\Pi)}_{\mu\nu}\,\delta g^{\mu\nu}\,\dd\Sigma.
          \label{eq:delta-E-Pi-2}
        \end{equation}
        Requiring~\eqref{eq:first-law} to hold with $\beta(x)^{-1}$ as in~\eqref{eq:beta-decomp},
        then matching the two sides at leading IR order requires
        \begin{equation}
          \frac{1}{16\pi G_N}\,G_{\mu\nu}
          = \beta(x)^{-1} \cdot \half\,T^{(\Pi)}_{\mu\nu},
          \label{eq:matching}
        \end{equation}
        which is consistent with the Einstein equation when $\beta(x)^{-1}$ is evaluated
        at the equilibrium value.
        In spectral units where $\beta_0^{-1} = 1$, the curvature correction
        $-\frac{1}{6}R$ is of order $R\ell_\chi^2$ and is absorbed into the subleading
        remainder (see Remark~\ref{rem:gauge}).
        The identity~\eqref{eq:first-law} is therefore a structural compatibility condition:
        it holds if and only if the geometry satisfies the Einstein equation at leading
        infrared order.
        This establishes~\eqref{eq:first-law} as a compatibility condition rather than
        a postulate.
    \end{proof}

    \begin{remark}[Nature of the identity]
      \label{rem:identity-nature}
      Equation~\eqref{eq:first-law} is not a definition of $\beta(x)$ from $\delta\Spi$
      and $\delta E_\Pi$: both sides are independently defined.
      $\beta(x)^{-1}$ is defined from the heat kernel via Definition~\ref{def:beta};
      $\delta E_\Pi$ is defined from the matter action via~\eqref{eq:delta-E-Pi}.
      The identity~\eqref{eq:first-law} is therefore a non-trivial constraint, not a
      tautology.
    \end{remark}
